---
title: "Introduction: fitting a multilevel hidden Markov model"
mainfont: Arial
params:
  answers: false
fontsize: 12pt
urlcolor: blue
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: paper
    pandoc_args: --output=Intro_pract.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mHMMbayes)
emotion_data <- readRDS("./data/data_Rowland2020.rds")
```

This is the first practical, where we introduce the dataset and we use it to fit a (2 state) multilevel hidden Markov model. 

You can use your preferred way of working in R to do the practicals. Our preferred way is this:

- Create a new folder with a good name, e.g., `practicals_mHMM` 
- Open RStudio
- Create a new project from RStudio, which you associate with the folder
- Create a `raw_data` subfolder
- Create an R script for the current practical, e.g., `introduction.R`
- Create your well-documented and [well-styled](https://style.tidyverse.org/) code in this R script

The answers to the practicals are available in the `*_answers.html` files. Try to work out the answer yourself (e.g., inspecting help files and examples in them provided in the `mHMMbayes` package, looking at the tutorial vignette available online) before looking at this!

In all practicals in this workshop, we make use of the `mHMMbayes` package, version 1.0.0. You can load the package like 

```{r, eval=FALSE}
library(mHMMbayes)
```

In this practical, we will also use the following packages:

```{r, eval=FALSE}
library(ggplot2)
```

## The data
We will use an open access dataset from Rowland and Wenzel (2020), available on OSF [`here`](https://osf.io/jmz2n/). The data consists of one-hundred twenty-five undergraduate students from the University of Mainz in Germany that completed a 40-day ambulatory assessment six times a day, reporting on their affective experiences *happy*, *excited*, *relaxed*, *satisfied*, *angry*, *anxious*, *depressed*, and *sad*. These items were scored with a visual analog slider from 0 to 100. Before the collection of the ESM measurements, participants of the study were randomly assigned to a group receiving a weekly mindfulness treatment during the ESM study and a control group. A cleaned version of the data kindly provided in the data archive as part of Haslbeck, Ryan, & Dablander (2023), the repository can be found [`here`](https://github.com/jmbh/EmotionTimeSeries), and the specific dataset we will be working with [`here`](https://github.com/jmbh/EmotionTimeSeries/tree/master/DataClean/Rowland2020). It is an `rds` file, which is a convenient, portable, and fast binary file format for R.

---

**Exercise 1**
Download the dataset and save it in a nice location, e.g., a `raw_data` folder inside your R project.

---

---

**Exercise 2**
Load the dataset in R using the function `readRDS()`. Give the dataset the name `emotion_data`. Then, inspect the first few rows of the data.

```{r, include = params$answers, eval=FALSE}
emotion_data <- readRDS("../../data/data_Rowland2020.rds")
```

```{r, include = params$answers}
# inspect the first few rows
head(emotion_data)
```
---


---

**Exercise 3**
Inspect the data over time using visualization. Most useful is visualizing time sequences for subjects separately, e.g., by using the option `facet_wrap(vars(subj_id))` in ggplot. 

```{r, include = params$answers}
# inspect the first few rows
library(ggplot2)

# restructure the data for ggplot
emotion_data$beep2 <- (emotion_data$dayno - 1) * 6 + emotion_data$beep

ggplot_emotion <- rbind(data.frame(ID = emotion_data$subj_id, 
                                 beep2 = emotion_data$beep2, 
                                 outcome = emotion_data$happy,
                                 emotion = "happy"),
                      data.frame(ID = emotion_data$subj_id, 
                                 beep2 = emotion_data$beep2, 
                                 outcome = emotion_data$excited,
                                 emotion = "excited"),
                      data.frame(ID = emotion_data$subj_id, 
                                 beep2 = emotion_data$beep2, 
                                 outcome = emotion_data$relaxed,
                                 emotion = "relaxed"),
                      data.frame(ID = emotion_data$subj_id, 
                                 beep2 = emotion_data$beep2, 
                                 outcome = emotion_data$satisfied,
                                 emotion = "satisfied"),
                      data.frame(ID = emotion_data$subj_id, 
                                 beep2 = emotion_data$beep2, 
                                 outcome = emotion_data$angry,
                                 emotion = "angry"),
                      data.frame(ID = emotion_data$subj_id, 
                                 beep2 = emotion_data$beep2, 
                                 outcome = emotion_data$anxious,
                                 emotion = "anxious"),
                      data.frame(ID = emotion_data$subj_id, 
                                 beep2 = emotion_data$beep2, 
                                 outcome = emotion_data$depressed,
                                 emotion = "depressed"),
                      data.frame(ID = emotion_data$subj_id, 
                                 beep2 = emotion_data$beep2, 
                                 outcome = emotion_data$sad,
                                 emotion = "sad"))

ggplot_emotion$emotion <- as.factor(ggplot_emotion$emotion)

ggplot(ggplot_emotion[ggplot_emotion$ID %in% c(1:5),]) +
  geom_line( mapping = aes(x = beep2, y = outcome, group = emotion, color = emotion)) +
  facet_wrap(vars(ID), ncol = 1)


```

---

## Setting up the multilevel hidden Markov model 

In this section, you will fit a 2-state multilevel hidden Markov model. For this, you need a `data.frame` in which only the affective variables you want to use in the model are included, plus the subject ID as the first column.  Please feel free to only use a subset of the provided affective variables, but do make sure to select at least 2 affective variables. 

---

**Exercise 4**
Create a dataset `emotion_mHMM` which has the subject ID variable in the first column, followed by only the affective variables you want to use in the hidden Markov model. 

```{r, include = params$answers}
# inspect the first few rows
emotion_mHMM <- data.frame(subj_ID = emotion_data$subj_id,
                           happy = emotion_data$happy,
                           excited = emotion_data$excited,
                           relaxed = emotion_data$relaxed,
                           satisfied = emotion_data$satisfied,
                           angry = emotion_data$angry,
                           anxious = emotion_data$anxious,
                           depressed = emotion_data$depressed,
                           sad = emotion_data$sad)
```

---

---

**Exercise 5**
Set up the first set of model input arguments: `m` for the number of states the model should infer, `n_dep` for the number of dependent variables used for fitting the model, and starting values for gamma and the emission distribution, saved in the objects `start_gamma` and `start_emiss`, respectively.

Note 1: `start_gamma` should be a `m` by `m` matrix and `start_emiss` should be a list with the number of elements equal to the the number of dependent variables used specified in `n_dep`. Each element within the list should be a matrix with 2 columns and `m` rows. The Normal emission distribution means are contained in the first column of this matrix, and the standard deviations of the state specific Normal emission distribution are contained int the second column of the matrix. 

Note 2: Make sure that the starting values are sensible. That is, for gamma, usually the elements contained on the diagonal (i.e., the self-transitions) are larger than the off-diagonal values. With respect to the emission distribution, if we assume a 2-state model, it is sensible to assume that we will have a 'positive emotion' state and a 'negative emotion' state. Define the means of the variable and state specific emission distributions to align with this idea. Do not use equivalent means for one dependent variable over the states! Regarding the standard deviations of the emission distribution, define them such that the emission distribution for state 1 and 2 together cover the entire range of observations (remember, participants could use a slider ranging from 0 to 100) with reasonable support. For example, given our data, it would not make sense to use a standard deviation of 3 (as 95% of the mass of the distribution will only span mean+/- 2 * 3 values out of 100), but a standard deviation of 10 or 15 would make sense. 

```{r, include = params$answers}
# setting up input parameters, part 1
m <- 2
n_dep <- 8

start_gamma <- matrix(c(0.7, 0.3, 
                        0.3, 0.7), byrow = TRUE, ncol = m, nrow = m)

start_emiss <- list(matrix(c(70, 15,                                     # happy
                             30, 15), byrow = TRUE, ncol = 2, nrow = m), 
                    matrix(c(70, 15,                                     # excited
                             30, 15), byrow = TRUE, ncol = 2, nrow = m), 
                    matrix(c(70, 15,                                     # relaxed
                             30, 15), byrow = TRUE, ncol = 2, nrow = m), 
                    matrix(c(70, 15,                                     # satisfied
                             30, 15), byrow = TRUE, ncol = 2, nrow = m), 
                    matrix(c(15, 10,                                     # angry
                             40, 15), byrow = TRUE, ncol = 2, nrow = m), 
                    matrix(c(15, 10,                                     # anxious
                             40, 15), byrow = TRUE, ncol = 2, nrow = m), 
                    matrix(c(15, 10,                                     # depressed
                             40, 15), byrow = TRUE, ncol = 2, nrow = m), 
                    matrix(c(15, 10,                                     # sad
                             40, 15), byrow = TRUE, ncol = 2, nrow = m))
```

---

---

**Exercise 6**
Set up a weakly informative prior for the continuous emission distribution using the function `prior_emiss_cont` and save the object in `emotion_prior_emiss`. Again, align the values of the hyper-prior means of the Normal distribution with your hypothesized structure of a 'positive emotion' state and a 'negative emotion' state. Do not use equavalent means over states within the same dependent variable. Tip: for an overview of the usage and an example, check the help file (`?prior_emiss_cont`). 

```{r, include = params$answers}
# specifying weakly informative prior for continuous emission distributions

emotion_prior_emiss <- prior_emiss_cont(
  gen = list(m = m, n_dep = n_dep),
  emiss_mu0 = list(matrix(c(70, 30), nrow = 1),  # happy
                   matrix(c(70, 30), nrow = 1),  # excited
                   matrix(c(70, 30), nrow = 1),  # relaxed
                   matrix(c(70, 30), nrow = 1),  # satisfied
                   matrix(c(15, 40), nrow = 1),  # angry
                   matrix(c(15, 40), nrow = 1),  # anxious
                   matrix(c(15, 40), nrow = 1),  # depressed
                   matrix(c(15, 40), nrow = 1)), # sad 
  emiss_K0 = rep(list(1), n_dep),
  emiss_V = rep(list(rep(5^2, m)), n_dep),
  emiss_nu = rep(list(1), n_dep),
  emiss_a0 = rep(list(rep(1.5, m)), n_dep),
  emiss_b0 = rep(list(rep(20, m)), n_dep),
)

```

---

## Fitting the multilevel hidden Markov model 

Now we will use all ingredients to fit a 2 state multilevel hidden Markov model. 

500 iterations ~ 3 minutes on reasonably fast laptop 


## Inspecting the output 

Inspect the global output by using the inbuilt print and summary options for the `mHMMbayes` package ()


## References

- J. Haslbeck, O. Ryan, & F. Dablander (2023). Multimodality and skewness in emotion time series. Emotion, 23(8), 2117–2141. DOI: 10.1037/emo0001218.
- Z. Rowland & M. Wenzel (2020). Mindfulness and affect-network density: Does mindfulness facilitate disengagement from affective experiences in daily life?. Mindfulness, 11(5), 1253-1266. DOI: 10.1007/s12671-020-01335-4
